Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-04-08 Ben_Manuscripts/hdphmm/Outline.tex
Some comments.

  % BJC5: For the paper, this paragraph and figure might be better suited for the supporting information
  % BJC5: maybe even for the thesis
  \noindent The states identified by the IHMM are heavily influenced by the Gaussian prior
  placed on $\mathbf{c}$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of $A$ and $\Sigma$ do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the mean level of 
    each state lies within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to lie halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the
    maximum and minimum to be 2 standard deviations from the prior mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

@@

  \noindent We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  	\item For each solute studied, we identified 200-325 independent states, each
   	with separate VAR(1) parameters.
   	\item Many of these states exhibit very similar dynamical behavior except their
   	mean levels are different.
  \end{itemize}

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached. 
  \begin{itemize}
   	\item We used the Ward linkage criteria, which works to minimize the sum of
   	the squared differences within all clusters.
   	\item We elected to choose the number of clusters rather than the distance
   	threshold.
   	\item For our data, non-parameteric methods such as Bayesian Gaussian mixture
   	models tend to delocalize the clusters in parameter space (see Supporting
   	Information).
   	%BJC: will show Gaussian mixture model which has a cluster engulfing another cluster, so it's tails are on both sides of the center cluster.
  \end{itemize}

@@

  % BJC5: not sure what to title this. Or break it into more sections.
  \subsection{Calculation of Physical Interactions}\label{method:interactions}

@@

  We estimated the lifetime of hydrogen bonds by recording the length of 
  sequential frames where solute remained hydrogen bonded.
  \begin{itemize}
    \item We still counted sequences where hydrogen bonds were broken for
    a single frame before reforming.
    \item We reported the 95th percentile of hydrogen bond lifetimes.
  \end{itemize}

@@

  Solutes in this system show a wide range of behavior influenced by the 
  heterogeneity of the membrane's nanostructure as well as the interactions 
  between monomer and solute chemical functionality. 
  \begin{itemize}  
  	\item Our application of the IHMM does a good job of identifying and
  	distinguishing these behaviors.   	
  	\item The following approach to analysis demonstrates how one can discover and explain the very complex behavior exhibited by solutes over long MD trajectories 
   in terms of solute-membrane interactions.
%  	\item Clustering the behaviors helps us to tell a coherent story which
%  	emphasizes those interactions which most heavily influence solute transport.
	\item Further, we can use the IHMM in order to generate stochastic trajectories,
	with dynamics similar to MD, so that we can estimate macroscopic transport
	properties.
  \end{itemize}

@@

  We applied the IHMM to all trajectories of each of the four solutes studied.
  \begin{itemize}
    \item Based on our choice of clustering, we identified over 30 distinct dynamical
    modes for each solute.
    \item Many of these modes are sampled very infrequently and thus do not merit significant study.
    \item In this section, we explore a subset of this data, focusing on the states
    and dominant interactions most common to solute trajectories. 
  \end{itemize}

  \noindent Methanol exhbits a range of dynamic behavior.
  \begin{itemize}
    \item In Figure~\ref{fig:mechanism_map}, we plot a representative 2D methanol 
    trajectory in cylindrical coordinates.
    \item The coordinates are colored according to the cluster with which methanol's
    dynamics are most consistent, which implies that this particular methanol 
    trajectory exhibits 8 distinct dynamical behaviors (in order of their first appearance,
    they are colored dark yellow, red, light blue, light green, orange, blue, teal and lime green). 
    \item The solute timeseries are coupled with color coded bars which describe the
    solute's local environment and physical interactions that it undergoes at that
    trajectory frame.
    \item From the top, they describe whether the solute is associated with sodium,
    whether the solute is hydrogen bonding with the monomer as well as to what part of
    the monomer it is hydrogen bonding, and finally, the local number density 
    surrounding the solute. 
    %and the radial location of the solute with respect to the nearest pore
    %center (this information is redundant but can be a useful visual aid).
  \end{itemize}

@@

  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}

  %BJC5: something we can do for the supporting information to show that this approach is valid: Eliminate nonidealities in our MD trajectories (compared to a pure VAR process) and use toy data to generate trajectories based on some number of states with different parameters. Generate an MSD with those. See if we can reproduce with our model. Can also use this to test new strategies.

  We can use the IHMM in order to generate stochastic trajectory realizations
  which bear some qualitatively similar characteristics to those output
  by our MD simulations.
  \begin{itemize}
    \item Because we clustered the data, we are left with a single transition
    probability matrix which allows transitions between dynamical behaviors
    shared by the 24 MD trajectories.
    \item In Figure~\ref{fig:trajectory_realizations_MET}, we show some stochastic
    realizations generated by our models compared directly to MD.
    \item The trajectories show clear periods of entrapment with intermittent hops.
  \end{itemize}

  %BJC5: the goal is to figure out how to fix this for the paper.
  %BJC5: this is more of a 'how can this be improved' section for the dissertation.
  The MD trajectories show much broader variability in their overall motion than realizations
  of the IHMM model. 
  \begin{itemize}
  		\item Some trajectories span a wide range of $z$ coordinates, while others stay
  		trapped for nearly the entire simulation.
  		\item Our model is very unlikely to generate trajectories with dwell times similar to 
  		the third MD trajectory shown in Figure~\ref{fig:trajectory_realizations_MET}.
        \item This is because the realizations generated by our model are representative
        of the average behavior of solutes studied.
        \item It is possible that certain states with long dwell times should have been
        clustered as their own, but the transition probabilities would have high uncertainties.
        \item It is also possible that we should not be using a static transition matrix in 
        order to generate stochastic trajectories, but rather sample the rows of the transition
        matrix from a Dirichlet distribution since average dwell times are highly sensitive to
        self transition probabilities.
  \end{itemize}
  %BJC5: maybe a figure showing the MSDs of all the MD trajectories versus a subset of 
  % stochastically generated trajectories. I image the stochastic ones will have a much
  % tighter distribution.

@@

  %BJC3: I still need to implement this into my mean first passage time code, but it will very closely follow the 
  % methodology of the last paper. However, if I see that the decay in flux is
  % near 2 (which I expect), then I may add a short section showing how you can
  % simply use the ratio of the MSD slopes in order to estimate selectivity.
  % I think they should give the same results. Haven't tested it yet though. 
  %BJC5: currently running these simulations then will write this section up.

>>>>

  % BJC5: For the paper, this paragraph and figure might be better suited for the supporting information
  % BJC5: maybe even for the thesis
  % MRS6: I agree on the details in SI, the thesis belongs in the paper. 
  \noindent The states identified by the IHMM are heavily influenced by the Gaussian prior
  placed on $\mathbf{c}$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of $A$ and $\Sigma$ do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the mean level of 
    each state lies within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to lie halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the
    maximum and minimum to be 2 standard deviations from the prior mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

@@

  \noindent We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  	\item For each solute studied, we identified 200-325 independent states, each
   	with separate VAR(1) parameters.
   	\item Many of these states exhibit very similar dynamical behavior except their
   	mean levels are different.
        %MRS6: should specify mean in both r and z if that's what you mean.
  \end{itemize}

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached. 
  \begin{itemize}
   	\item We used the Ward linkage criteria, which works to minimize the sum of
   	the squared differences within all clusters.
        %MRS6: explain why Ward (make the cluster as minimally disperse?)
   	\item We elected to choose the number of clusters rather than th distance
   	threshold.
   	\item For our data, non-parameteric methods such as Bayesian Gaussian mixture
   	models tend to delocalize the clusters in parameter space (see Supporting
   	Information).
   	%BJC: will show Gaussian mixture model which has a cluster engulfing another cluster, so it's tails are on both sides of the center cluster.
  \end{itemize}

@@

  % BJC5: not sure what to title this. Or break it into more sections.
  % MRS6: ``exploring mechanism?''
  \subsection{Calculation of Physical Interactions}\label{method:interactions}

@@

  We estimated the lifetime of hydrogen bonds by recording the length of 
  sequential frames where solute remained hydrogen bonded.
  \begin{itemize}
    \item We still counted sequences where hydrogen bonds were broken for
    a single frame before reforming.
    \item We reported the 95th percentile of hydrogen bond lifetimes.
      %MRS6: why 95th percentile?
  \end{itemize}

@@

  Solutes in this system show a wide range of behavior influenced by the 
  heterogeneity of the membrane's nanostructure as well as the interactions 
  between monomer and solute chemical functionality. 
  \begin{itemize}  
  	\item Our application of the IHMM does  good job of identifying and
  	distinguishing these behaviors.   	
        %MRS6: will need to quantify ``good job''
  	\item The following approach to analysis demonstrates how one can discover and explain the very complex behavior exhibited by solutes over long MD trajectories 
   in terms of solute-membrane interactions.
%  	\item Clustering the behaviors helps us to tell a coherent story which
%  	emphasizes those interactions which most heavily influence solute transport.
	\item Further, we can use the IHMM in order to generate stochastic trajectories,
	with dynamics similar to MD, so that we can estimate macroscopic transport
	properties.
  \end{itemize}

@@

  We applied the IHMM to all trajectories of each of the four solutes studied.
  \begin{itemize}
    \item Based on our choice of clustering, we identified over 30 distinct dynamical
    modes for each solute.
    \item Many of these modes are sampled very infrequently and thus do not merit significant study.
      %MRS6: explain why don't merit additional study - should relate to how does not affect final results.
    \item In this section, we explore a subset of this data, focusing on the states
    and dominant interactions most common to solute trajectories. 
  \end{itemize}

  %MRS6: likely just the summary of the methanol behavior should go in the main paper - the rest should be SI.
  \noindent Methanol exhbits a range of dynamic behavior.
  \begin{itemize}
    \item In Figure~\ref{fig:mechanism_map}, we plot a representative 2D methanol 
    trajectory in cylindrical coordinates.
    \item The coordinates are colored according to the cluster with which methanol's
    dynamics are most consistent, which implies that this particular methanol 
    trajectory exhibits 8 distinct dynamical behaviors (in order of their first appearance,
    they are colored dark yellow, red, light blue, light green, orange, blue, teal and lime green). 
    \item The solute timeseries are coupled with color coded bars which describe the
    solute's local environment and physical interactions that it undergoes at that
    trajectory frame.
    \item From the top, they describe whether the solute is associated with sodium,
    whether the solute is hydrogen bonding with the monomer as well as to what part of
    the monomer it is hydrogen bonding, and finally, the local number density 
    surrounding the solute. 
    %and the radial location of the solute with respect to the nearest pore
    %center (this information is redundant but can be a useful visual aid).
  \end{itemize}

@@

  % MRS6: ideal for the paper would be summary infoormation about the 4 molecules, with more complete data from which the summary is drawn in SI.
  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}

  %BJC5: something we can do for the supporting information to show that this approach is valid: Eliminate nonidealities in our MD trajectories (compared to a pure VAR process) and use toy data to generate trajectories based on some number of states with different parameters. Generate an MSD with those. See if we can reproduce with our model. Can also use this to test new strategies.
  %MRS6: toy data is actually a VERY good idea.  USe a VAR process that looks like MD, and then try to cluster it, and figure out why different.  
  We can use the IHMM in order to generate stochastic trajectory realizations
  which bear some qualitatively similar characteristics to those output
  by our MD simulations.
  \begin{itemize}
    \item Because we clustered the data, we are left with a single transition
    probability matrix which allows transitions between dynamical behaviors
    shared by the 24 MD trajectories.
    \item In Figure~\ref{fig:trajectory_realizations_MET}, we show some stochastic
    realizations generated by our models compared directly to MD.
    \item The trajectories show clear periods of entrapment with intermittent hops.
  \end{itemize}

  %BJC5: the goal is to figure out how to fix this for the paper.
  %BJC5: this is more of a 'how can this be improved' section for the dissertation.
  %MRS6: got it.
  The MD trajectories show much broader variability in their overall motion than realizations
  of the IHMM model. 
  %MRS6: definitely an important poiint for the paper: clustering things togethere might make trajectories too typical.
  \begin{itemize}
  		\item Some trajectories span a wide range of $z$ coordinates, while others stay
  		trapped for nearly the entire simulation.
  		\item Our model is very unlikely to generate trajectories with dwell times similar to 
  		the third MD trajectory shown in Figure~\ref{fig:trajectory_realizations_MET}.
        \item This is because the realizations generated by our model are representative
        of the average behavior of solutes studied.
        \item It is possible that certain states with long dwell times should have been
        clustered as their own, but the transition probabilities would have high uncertainties.
        %MRS6: is it necessarily a problem that the transition probabiblities have high uncertainties? If one is drawing from the Dirichlet distribution, uncertainy isn't a big deal. The trapping just doesn't quite seem to be as long as MD.
        %MRS6: the other issue is apparent surpression on variance in the jumps, but that may be harder to get. 
        \item It is also possible that we should not be using a static transition matrix in 
        order to generate stochastic trajectories, but rather sample the rows of the transition
        matrix from a Dirichlet distribution since average dwell times are highly sensitive to
        self transition probabilities.
        %MRS6: right uncertainty matters less here.
  \end{itemize}
  %BJC5: maybe a figure showing the MSDs of all the MD trajectories versus a subset of 
  % stochastically generated trajectories. I image the stochastic ones will have a much
  % tighter distribution.

@@

  %BJC3: I still need to implement this into my mean first passage time code, but it will very closely follow the 
  % methodology of the last paper. However, if I see that the decay in flux is
  % near 2 (which I expect), then I may add a short section showing how you can
  % simply use the ratio of the MSD slopes in order to estimate selectivity.
  % I think they should give the same results. Haven't tested it yet though. 
  %BJC5: currently running these simulations then will write this section up. 
  %MRS6: it's possible that this section doesn't need to be in the paper, since it was describbed in the other one, and getting the selectivity isn't the main goal of this paper.
